---
title: "Grad Assignment"
author: "Ellie Grace Moore"
date: "4/18/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Hey causal cats! Welcome back to my blog. Today we will choose a specific causal assumption and then generate data that *violates* this assumption.

Choose one "Causal Inference Assumption" and write a blog post describing:

What the assumption is in words
What the assumption means mathematically
An example where the assumption is violated
Using the simulation techniques we've used in class (for example the "Lucy Land" simulations") Come up with a data set that violates the assumption you've chosen. Attempt to estimate a causal effect using this dataset and explain how wrong (or not) the final result is.


```{r}
set.seed(1)
weefle <- tibble(
    Z = rnorm(1000, 0, 1), #Confounder
    x_lin = 3 + Z + rnorm(1000, 0, 1), 
    x_p = exp(x_lin)/(1+exp(x_lin)))  #Exposure Probabilities
weefle$x_p[seq(1,1000, by = 4)] = 0 #Assigning some probabilities as 0

weefle <- weefle %>% mutate(
  x = rbinom(1000, 1, x_p) #Exposure assignments
)

d_ellie <- weefle %>%
  mutate(y0 = Z + x,
          y1 = Z + x,
          y_obs = ifelse(x == 1, y1, y0))
```


```{r}
d_ellie %>%
  summarise(true_causal_effect = mean(y1) - mean(y0))

d_ellie %>%
  summarise(observed_causal_effect = 
              sum(y_obs * x) / sum(x) -
              sum(y_obs * (1 - x)) / sum(1 - x))
```

```{r}
p <- glm(x ~ Z, data = d_ellie, family = binomial()) %>%
  predict(type = "response")
d_ellie <- d_ellie %>%
  mutate(prob_of_exposure = p)
```



```{r}
d_ellie %>%
  mutate(w_ate = x / p + (1 - x) / (1 - p)) %>%
  summarise(observed_causal_effect = 
              sum(y_obs * x * w_ate) / sum(x * w_ate) -
              sum(y_obs * (1 - x) * w_ate) / sum((1 - x) * w_ate))
```



